A thermodynamic view of dusty protoplanetary disks

### 2017-05-02 paper.tex
misc edits from sect 4 to sect 8.3

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks I:
axisymmetric stability}
 % \title{Dust-free modeling of dusty protoplanetary disks I:
%axisymmetric stability}
\author{Min-Kai Lin\altaffilmark{1,2,3}}
\author{Andrew N. Youdin\altaffilmark{1}}
%\affil{Department of Astronomy and Steward Observatory, 
%University of
 % Arizona, Tucson, AZ 85721, USA\\
%Institute of Astronomy and Astrophysics, Academia Sinica, 
%Taipei 10617, Taiwan}
\altaffiltext{1}{Department of Astronomy and Steward Observatory, 
University of
Arizona, Tucson, AZ 85721, USA}
\altaffiltext{2}{Steward Theory Fellow}
\altaffiltext{3}{Institute of Astronomy and Astrophysics, Academia Sinica, 
Taipei 10617, Taiwan
}
\email{mklin@asiaa.sinica.edu.tw}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled dust cannot cause axisymmetric 
 instabilties, however thin the dust layer is; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading generally stabilizes the vertical shear
  instability, because dust-gas coupling introduces buoyancy. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
correspondence in finding dusty analogs of other hydrodynamic
phenomena (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and to simulate dusty protoplanetary disks
using purely hydrodynamic codes. %We briefly describe extending this
%equivalence principle to polytropic and adiabatic gas.

>>>>

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks}
 % \title{Dust-free modeling of dusty protoplanetary disks I:
%axisymmetric stability}
\author{Min-Kai Lin\altaffilmark{1,2,3}}
\author{Andrew N. Youdin\altaffilmark{1}}
%\affil{Department of Astronomy and Steward Observatory, 
%University of
 % Arizona, Tucson, AZ 85721, USA\\
%Institute of Astronomy and Astrophysics, Academia Sinica, 
%Taipei 10617, Taiwan}
\altaffiltext{1}{Department of Astronomy and Steward Observatory, 
University of
Arizona, Tucson, AZ 85721, USA}
\altaffiltext{2}{Steward Theory Fellow}
\altaffiltext{3}{Institute of Astronomy and Astrophysics, Academia Sinica, 
Taipei 10617, Taiwan
}
\email{mklin@asiaa.sinica.edu.tw}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to entropy conservation
  following an adiabatic gas.  
  Motivated by this similarity, we develop a thermodynamical
  framework to study dusty gas dynamics in protoplanetary disks. We  
 show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled, thin dust layers cannot cause axisymmetric 
 instabilties; 
\item 
  the streaming instability neccessarily involves a pressure response
  that lags dust density; 
\item dust-loading introduce buoyancy forces that generally 
  stabilizes the vertical shear 
  instability that arises from a radial temperature gradient. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss future applications of the dusty/adiabatic gas
correspondence in finding dusty analogs of other hydrodynamic
phenomena (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and to simulate dusty protoplanetary disks
using purely hydrodynamic codes. %We briefly describe extending this
%equivalence principle to polytropic and adiabatic gas.

### 2017-05-02 result.tex
misc edits from sect 4 to sect 8.3

We present numerical solutions for vertically stratified,
non-self-gravitating disks assuming perfectly coupled dust
($\tstop=0$) so that the equilibria defined in \S\ref{eqm} 
are exact steady states.

Except in \S\ref{varHd}---\ref{vert_mixed}, we consider constant midplane dust-to-gas
ratios $\epsilon_0$ and characteristic thickness $H_\epsilon$. Then
the dust-to-gas ratio $\epsilon=\epsilon(z)$. In this limit any
growing modes must be associated with the imposed temperature gradient
(\S\ref{limits}). We are thus examining the effect of dust-loading on
the VSI, previously studied in pure gas disks
\citep[][\citetalias{lin15} in this section]{lin15}.

The parameters for this linear problem includes $\epsilon_0$, the dust
layer thickness $\Hd(r_0)$, and the perturbation radial wavenumber
$k_x$. We choose a power-law midplane gas density profile,
$\rho_\mathrm{g0}\propto r^p$ with $p = -1.5$. The fiducial power-law
index for the temperature profile is $q=-1$; and we set the gas disk
aspect-ratio $h_\mathrm{g}=0.05$. These values were also used by  
\citetalias{lin15}.

Two boundary conditions are needed for the $\tstop=0$
problem \citep[e.g.][]{lubow93}. We impose solid boundaries so that
$\delta v_z(\pm\zmax)=0$. We solve the linearized equations as a
generalized eigenvalue problem using a pseudo-spectral code adapted
from \citetalias{lin15}. Amplitudes are expanded in Chebyshev
polynomials up to order $N_z=512$. We check results using
Eq. \ref{vsi_check}.

\subsection{Qualitative expectations}\label{vsi_est}
\citetalias{lin15} found the appropriate way to compare 
destabilizing vertical shear to stabilizing vertical buoyancy is
$r\p_z\Omega/\OmK$ against $N_z^2/\OmK^2$. We see from \S\ref{vertshear}
and \S\ref{vbuoyancy} that $|r\p_z\Omega|\sim q h_\mathrm{g}\OmK$ for a
thin, gas dominated disk; while $N_z^2\sim \epsilon\OmK^2$. Thus we
expect dust-induced buoyancy forces to stabilize the disk against the
VSI where $\epsilon \gtrsim h_\mathrm{g}$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixHd} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy
    (dotted) in a locally isothermal, dusty disl with midplane dust-to-gas ratio
    of $\epsilon_0=10^{-3}$ (black) and $\epsilon_0=1$ (blue). 
    The dust layer thickness is fixed to $\Hd=0.99\Hg$ so the 
    dust-to-gas ratio $\epsilon$ is approximately constant with
    height. 
    \label{compare_vshear_fixHd}
    }
\end{figure}

Fig. \ref{vsi_dust_loading} show unstable modes for different values
of $\epsilon_0$ with fixed perturbation wavenumber  $k_x\Hg = 30$. The
eigenvalue distributions for $\epsilon_0 \leq 10^{-2}$ are similar to the
dust-free fiducial case considered by \citetalias{lin15}. This is
expected since $\epsilon_0 < \hgas$ (\S\ref{vsi_est}). Eigenvalues 
consists 
of the roughly horizontal `body modes', and the nearly-vertical
`surface modes' \citep[which are associated with the imposed vertical
boundaries][]{barker15}.

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably, surface modes, which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$ (i.e. $\epsilon_0> \hgas$).   
The body modes' growth rates remain $ O(h_\mathrm{g}\OmK)$ 
but their oscillation frequency increases with
$\epsilon_0$. %i.e. increases with buoyancing freq 
The total number of modes do not change
significantly. This is in contrast with the effect of increasing
cooling times, where \citetalias{lin15} find fewer unstable modes.
%until the system is eventually completely stable.

@@

In Fig. \ref{vsi_dust_loading_vareps} we plot the growth rates as a
function of $\epsilon_0$ for different perturbation wavenumbers
$k_x$. We find dust-loading stabilizes the VSI more effectively for shorter
wavelength perturbations. This is because for high wavenumbers the
dominant modes are surface modes, which are effectively stabilized by
dust-loading since buoyancy forces are larges near the disk
boundaries. The figure suggest that VSI becomes much less efficient
for $\epsilon_0\gtrsim 0.1$ and $k_x\Hg\gtrsim 50$.

@@

\subsection{Effect of dust thickness} 
We now vary the dust layer thickness $\Hd$ but fix the metalicity 
$Z \equiv \epsilon_0 \Hd/\Hg = 0.03$. This then gives the
required midplane dust-to-gas ratio. Since we will consider thin dust
layers, here we use a smaller 
domain with $\zmax=2\Hg$ so that $\epsilon$ does not become
too small.

@@

Fig. \ref{result2d_fixZ} compares the fastest growing VSI body modes  
with $k_x\Hg=30$ for the two cases above.\citepalias[The thinner domain
  adopted here eliminates surface modes, ][]{lin15}.   
We find very similar mode 
frequencies 
\begin{align*}
  \sigma = \begin{cases}
    \left(0.3053\ii - 0.8142\right)h_\mathrm{g}\OmK & \Hd=0.99\Hg, \\
    \left(0.3178\ii - 1.2237\right)h_\mathrm{g}\OmK & \Hd=0.1\Hg,
  \end{cases}
\end{align*}
since the vertical shear profile is similar throughout most of the
disk. However, meridional motions are surpressed near the midplane of
the $\Hd=0.1\Hg$ disk, as expected from the larger buoyancy frequency
as compared to vertical shear there. This leads to a structure
analogous to PPD dead zones: a quiescent midplane between active
surface layers \citep{gammie96}.

@@

Fig. \ref{compare_eigenvals_fixZ} shows the maximum VSI growth rates
as a function of $\Hd$. As before, we find growth
rates are most affected by the vertical structure of the dust layer
when the perturbation wavenumer is large. Notice VSI growth rates converge as
$\Hd\to 0$. %to dust free values?    
Thus a thin dust layer, however large its associated vertical shear,
does not affect the VSI. The non-monotonic behavior for $\Hd\gtrsim
0.5\Hg$ arises because the vertical buoyancy frequency 
\begin{align*}
N_z^2(H_d;z,Z) \simeq &Z\Hg z^2
\exp{\left(-\frac{z^2}{2\Hg^2}\right)}\OmK^2\notag\\
&\times 
\frac{1}{\Hd}\left(\frac{1}{\Hd^2} -
\frac{1}{\Hg^2}\right)\exp{\left(-\frac{z^2}{2\Hd^2}\right)}  
\end{align*}
is a non-monotonic function of $\Hd$ at fixed $z$. At $z=\Hg$ and $z=2\Hg$
the buoyancy frequency is maximized for $\Hd\simeq0.5\Hg$ and
$\Hd\simeq 0.8\Hg$, respectively. This is consistent with the abscissa
of minima in growth rates in Fig. \ref{compare_eigenvals_fixZ}.

@@

\subsection{Axisymmetric stability of ultra-thin dust layers}
In \S\ref{iso_perfect} we asserted that in the limit of perfectly
coupled dust: radially smooth, strictly
isothermal disks are stable against axisymmetric perturbations, no
matter how thin the dust layer is. We now demonstrate this
numerically.

For comparison with previous studies of dusty disks, we also use 
the Richardson number $\rich \equiv N_z^2/\left(r\p_z\Omega\right)^2$
to label the calculations here \citep{youdin02}.   
Numerical simulations show    
non-axisymmetric instabilities develops when $\rich\lesssim0.1$, brought about
by very thin dust layers \citep{chiang08, lee10}. We show that \emph{axisymmetric}
instabilities never develop, however small $\rich$.

We shall consider ultra thin dust layers with $\Hd\leq 0.01\Hg$ and thus
restrict the vertical domain to $\zmax = 0.02\Hg$. We fix the 
metalicity $Z=0.01$ so the midpane dust-to-gas ratio $\epsilon_0$ is 
$O(1)$.  We consider radial wavenumbers such $k_x\Hd=1$.

Fig. \ref{ultra_thin} shows the maximum growth rate  as a function of
the radial temperature gradient, $q$. For all cases the vertical
shear is dominated by that due to the dust layer (see
\S\ref{vertshear}). However, we see that growth rates $s\propto
|q|$, i.e .growth rates vanish if the disk was strictly isothermal. 
In particular, this holds for $\rich < 0.1$, the critical value
for non-axisymmetric instabilities.

Axisymmetric instability is associated with the thermal
contribution to vertical shear: as $q\to0$,  $\p_z\Omega$ becomes
entirely due to dust-settling and there is no instability ($s\to
0$). This result is independent of $\rich$, so the Richardson number
is does not characterize the axisymmetric stability of dust
layers.

@@

Here we allow the dust-to-gas ratio $\epsilon$ to depend on
$r$. Specifically we let $\epsilon_0\propto r^{-1}$ and
$H_\epsilon \propto \Hg$ (cf. constant values in the previous
calculations). Then
\begin{align*}
  \frac{\p\epsilon}{\p r} =
  \left(\frac{z^2}{H_\epsilon^2}\frac{d\ln{\Hg}}{dr} -
    \frac{1}{r}\right)\epsilon.   
\end{align*}
We fix $Z=0.01$, $\Hd=0.8\Hg$ at $r=r_0$, and
$k_x\Hg=30$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_varHd} 
  \caption{Unstable VSI modes in the disk models of
    Fig. \protect\ref{compare_vshear_varHd}.  
    \label{vsi_dust_varHd}
    }
\end{figure}

@@

In \S\ref{vert_mixed_theory} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$. To demonstate this numerically we consider the
same disk as above but with $q=0$ and choose $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. Then vertical shear $\p_z\Omega$ is attributed
to the radially-varying dust-to-gas ratio (see Eq. \ref{vshear2}).

We show unstable modes for this disk in
Fig. \ref{vert_mixed_modes}. In order to obtain appreciable growth  
rates, we needed to consider super-Solar metalicity $Z=0.03$ and high
radial wavenumber $k_x\Hg=1800$. As expected from the discussion in
\S\ref{vert_mixed_theory}, the disk admits purely growing modes with
$\omega=0$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/vert_mixed_modes} 
  \caption{Unstable modes in a strictly isothermal disk ($q=0$) with
    vertically uniform dust-to-gas ratio. The instability is due to
    vertical shear $\p_z\Omega\neq0$ arising from radial variations in
    the dust-to-gas ratio, $d\epsilon/dr\neq 0$. \label{vert_mixed_modes}
    }
\end{figure}

\subsection{Validity}
We considered $\tstop=0$ in order to perform stability analysis on an
exact steady state. When $\tstop\neq0$, a stratified dusty 
disk initialized as in \S\ref{eqm} would evolve 
%due to dust-gas relative drift. In particular, in the
%absence of turbulent/diffusive processes, dust particles 
as dust settles to the
midplane on a timescale $t_\mathrm{settle}\sim 1/\tstop\OmK^2$
\citep{takeuchi02}. However, we expect the above results to remain
valid provided the VSI growth
timescale $t_\mathrm{grow}\ll t_\mathrm{settle}$. Since 
VSI growth rates are $O(h_\mathrm{g}\OmK)$, we require 
\begin{align}
  \tstop\OmK \ll h_\mathrm{g}. 
\end{align} 
For thin PPDs with $h_\mathrm{g}\sim 0.05$, this is satisfied 
for $\tstop\OmK \ll O(10^{-2})$.

>>>>

We now present numerical solutions for vertically stratified,
non-self-gravitating disks assuming perfectly coupled dust
($\tstop=0$) so the equilibria
defined in \S\ref{eqm} are exact steady states.

Apart from \S\ref{varHd}---\ref{vert_mixed}, we consider constant
midplane dust-to-gas ratios $\epsilon_0$ and characteristic thickness
$H_\epsilon$. Then the dust-to-gas ratio $\epsilon=\epsilon(z)$. In 
this limit any growing modes must be associated with the imposed temperature gradient
(\S\ref{limits}). We are thus examining the effect of dust-loading on
the VSI previously studied in pure gas disks 
\citep[][\citetalias{lin15} in this section]{lin15}.

The parameters for this linear problem includes $\epsilon_0$, the dust
layer thickness $\Hd(r_0)$, and the perturbation radial wavenumber
$k_x$. We choose a midplane gas density profile 
$\rho_\mathrm{g0}\propto r^{-3/2}$. The fiducial power-law 
index for the temperature profile is $q=-1$; and we set the gas disk
aspect-ratio $h_\mathrm{g}=0.05$. These values were also used by  
\citetalias{lin15}.

%Two boundary conditions are needed for the $\tstop=0$
%problem \citep[e.g.][]{lubow93}. 
We impose solid vertical boundaries so that $\delta
v_z(\pm\zmax)=0$. We solve the linearized equations as a generalized
eigenvalue problem using a pseudo-spectral code adapted  
from \citetalias{lin15}. Amplitudes are expanded in Chebyshev
polynomials up to order $N_z=512$. %We check results using
%Eq. \ref{vsi_check}.

\subsection{Qualitative expectations}\label{vsi_est}
\citetalias{lin15} found the appropriate way to compare 
vertical shear (destabilizing) and  vertical buoyancy (stabilizing)
is $r\p_z\Omega/\OmK$  against $N_z^2/\OmK^2$. From 
\S\ref{vertshear} and \S\ref{vbuoyancy} we find $|r\p_z\Omega|\sim q
h_\mathrm{g}\OmK$ for a 
thin, gas dominated disk; while $N_z^2\sim \epsilon\OmK^2$. Thus we
expect dust-induced buoyancy forces to stabilize the disk against the
VSI where $\epsilon \gtrsim h_\mathrm{g}$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_vshear_Nz2_fixHd} 
  \caption{Vertical shear rate (solid) compared to vertical buoyancy
    (dotted) in a locally isothermal, dusty disk with midplane dust-to-gas ratio
    of $\epsilon_0=10^{-3}$ (black) and $\epsilon_0=1$ (blue). 
    The dust layer thickness is fixed to $\Hd=0.99\Hg$ so the 
    dust-to-gas ratio $\epsilon$ is approximately constant with
    height. 
    \label{compare_vshear_fixHd}
    }
\end{figure}

Fig. \ref{vsi_dust_loading} show unstable modes for different values
of $\epsilon_0$ with fixed perturbation wavenumber  $k_x\Hg = 30$. The
eigenvalue distributions for $\epsilon_0 \leq 10^{-2}$ are similar to the
dust-free fiducial case considered by \citetalias{lin15}. This is
expected since $\epsilon_0 < \hgas$ (\S\ref{vsi_est}). Eigenvalues 
consists 
of the roughly horizontal `body modes', and the nearly-vertical
`surface modes' \citep[which are associated with the imposed vertical
boundaries, ][]{barker15}.

We find that increasing the dust-to-gas ratio reduce VSI growth
rates. Notably, surface modes, which are typically fastest growing in
the dust-free case, are surpressed in dusty disks for $\epsilon_0\geq
0.1$ (i.e. $\epsilon_0> \hgas$).   
The body modes' growth rates remain $ O(h_\mathrm{g}\OmK)$ 
but their oscillation frequency increases with
$\epsilon_0$. i.e. it increases with the vertical buoyancy. 
The total number of modes do not change 
significantly. This is in contrast with the effect of increasing
cooling times, for which \citetalias{lin15} find fewer unstable modes.
%until the system is eventually completely stable.

@@

In Fig. \ref{vsi_dust_loading_vareps} we plot the growth rates as a
function of $\epsilon_0$ for different perturbation wavenumbers
$k_x$. Dust-loading stabilizes the VSI more effectively for shorter
wavelength perturbations. This is because for high wavenumbers the
dominant modes are surface modes, which are effectively stabilized by
dust-loading as buoyancy forces are largest near the vertical 
boundaries. The figure suggest that VSI becomes much less efficient
for $\epsilon_0\gtrsim 0.1$ and $k_x\Hg\gtrsim 50$.

@@

\subsection{Effect of dust layer thickness} 
We now vary $\Hd$ but fix the metalicity 
$Z \equiv \epsilon_0 \Hd/\Hg = 0.03$ to obtain $\epsilon_0$. Since we
will consider thin dust layers, here we use a smaller   
domain with $\zmax=2\Hg$ so that $\epsilon$ does not become
too small.

@@

Fig. \ref{result2d_fixZ} compares the fastest growing VSI body modes  
with $k_x\Hg=30$ for the two cases above.\citepalias[The thinner domain
  adopted here eliminates surface modes, ][]{lin15}.   
We find very similar mode 
frequencies 
\begin{align*}
  \sigma = \begin{cases}
    \left(0.3053\ii - 0.8142\right)h_\mathrm{g}\OmK & \Hd=0.99\Hg, \\
    \left(0.3178\ii - 1.2237\right)h_\mathrm{g}\OmK & \Hd=0.1\Hg,
  \end{cases}
\end{align*}
since the vertical shear profile is similar throughout most of the
disk. However, meridional motions are surpressed near the midplane of
the $\Hd=0.1\Hg$ disk, as expected from the larger buoyancy frequency
relative to vertical shear there. This leads to a structure
analogous to PPD dead zones: a quiescent midplane between active
surface layers \citep{gammie96}.

@@

Fig. \ref{compare_eigenvals_fixZ} shows the maximum VSI growth rates
as a function of $\Hd$. As before, we find growth
rates are most affected by the vertical structure of the dust layer
when the perturbation wavenumer is large. Notice VSI growth rates converge as
$\Hd\to 0$. %to dust free values?    
Thus a thin dust layer, however large its associated vertical shear,
does not affect VSI growth rates. The non-monotonic behavior for $\Hd\gtrsim
0.5\Hg$ arises because the vertical buoyancy frequency 
\begin{align*}
N_z^2(H_d;z,Z) \simeq &Z\Hg z^2
\exp{\left(-\frac{z^2}{2\Hg^2}\right)}\OmK^2\notag\\
&\times 
\frac{1}{\Hd}\left(\frac{1}{\Hd^2} -
\frac{1}{\Hg^2}\right)\exp{\left(-\frac{z^2}{2\Hd^2}\right)}  
\end{align*}
is a non-monotonic function of $\Hd$ at fixed $z$. At $z=\Hg$ and $z=2\Hg$
the buoyancy frequency is maximized for $\Hd\simeq0.5\Hg$ and
$\Hd\simeq 0.8\Hg$, respectively. This is consistent with the abscissa
of minima in growth rates in Fig. \ref{compare_eigenvals_fixZ}.

@@

\subsection{Axisymmetric stability of ultra-thin dust layers}
The discussion in \S\ref{iso_perfect} imply strictly  
isothermal disks with a radially uniform dust-to-gas ratio are
stable against axisymmetric perturbations, no matter how thin the dust
layer is. We now demonstrate this numerically.

To connect with previous studies, here  we also use 
the Richardson number $\rich \equiv N_z^2/\left(r\p_z\Omega\right)^2$
to label calculations \citep{youdin02}.   
Numerical simulations show    
non-axisymmetric instabilities develops when $\rich\lesssim0.1$, brought about
by very thin dust layers \citep{chiang08, lee10}. We show that \emph{axisymmetric}
instabilities never develop, however small $\rich$.

We shall consider ultra thin dust layers with $\Hd\leq 0.01\Hg$ and thus
restrict the vertical domain to $\zmax = 0.02\Hg$. We fix the 
metalicity $Z=0.01$ so the midpane dust-to-gas ratio $\epsilon_0$ is 
$O(1)$.  We consider radial wavenumbers with $k_x\Hd=1$.

Fig. \ref{ultra_thin} shows the maximum growth rate  as a function of
the radial temperature gradient, $q$. For all cases the vertical
shear is dominated by that due to the dust layer (see
\S\ref{vertshear}). However, we see that growth rates $s\propto
|q|$, i.e .growth rates vanish in the strictly isothermal limit.  
In particular, this holds for $\rich < 0.1$, the critical value 
for non-axisymmetric instabilities.

Axisymmetric instability here is associated with the thermal
contribution to vertical shear: as $q\to0$,  $\p_z\Omega$ becomes
entirely due to dust-settling and there is no instability ($s\to
0$). This result is independent of $\rich$, so the Richardson number
is does not characterize the axisymmetric stability of dust
layers.

@@

We now consider a radially-varying dust-to-gas ratio. Specifically we
let $\epsilon_0\propto r^{-1}$ and $H_\epsilon \propto \Hg$
(cf. constant values in the previous calculations). Then
\begin{align*} 
  \frac{\p\epsilon}{\p r} =
  \left(\frac{z^2}{H_\epsilon^2}\frac{d\ln{\Hg}}{dr} -
    \frac{1}{r}\right)\epsilon.   
\end{align*}
Here we fix $Z=0.01$, $\Hd=0.8\Hg$, and $k_x\Hg=30$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_varHd} 
  \caption{Unstable VSI modes in the disk models of
    Fig. \protect\ref{compare_vshear_varHd}.  Black (blue) diamonds
    are mode frequencies for a disk with radially uniform (varying)
    dust-to-gas ratio. 
    \label{vsi_dust_varHd}
    }
\end{figure}

@@

In \S\ref{vert_mixed_theory} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$.  
To demonstate this numerically we consider the
same disk as above but with $q=0$ and choose $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. Then vertical shear $\p_z\Omega$ is then
attributed to the radially-varying dust-to-gas ratio (see
Eq. \ref{vshear2}).

We show unstable modes for this disk in
Fig. \ref{vert_mixed_modes}. In order to obtain appreciable growth   
rates, we needed to consider super-Solar metalicity $Z=0.03$ and high
radial wavenumber $k_x\Hg=1800$. As expected from the discussion in
\S\ref{vert_mixed_theory}, the disk admits purely growing modes with
$\omega=0$. This is distinct from the growing oscillations associated
with classic VSI discussed above.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/vert_mixed_modes} 
  \caption{Purely growing modes in a strictly isothermal disk ($q=0$) with
    vertically uniform dust-to-gas ratio. The instability is due to 
    vertical shear $\p_z\Omega\neq0$ arising from radial variations in
    the dust-to-gas ratio, $d\epsilon/dr\neq 0$. \label{vert_mixed_modes}
    }
\end{figure}

\subsection{Validity}
{\bf maybe incorporate into intro to this section?}
We considered $\tstop=0$ in order to perform stability analysis on an
exact steady state. When $\tstop\neq0$, a stratified dusty 
disk initialized as in \S\ref{eqm} would evolve 
%due to dust-gas relative drift. In particular, in the
%absence of turbulent/diffusive processes, dust particles 
as dust settles to the
midplane on a timescale $t_\mathrm{settle}\sim 1/\tstop\OmK^2$
\citep{takeuchi02}. However, we expect the above results to remain
valid provided the VSI growth
timescale $t_\mathrm{grow}\ll t_\mathrm{settle}$. Since 
VSI growth rates are $O(h_\mathrm{g}\OmK)$, we require 
\begin{align}
  \tstop\OmK \ll h_\mathrm{g}. 
\end{align} 
For thin PPDs with $h_\mathrm{g}\sim 0.05$, this is satisfied 
for $\tstop\OmK \ll O(10^{-2})$.
